Supplementary materials for 'The collapse of eastern Bering Sea snow crab

### 2023-07-25 manuscript_supmat.Rmd
revised tables

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.6035 & 0.0439 & 13.7531 & $<$ 0.0001 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(temperature) & 2.0477 & 2.4391 & 3.6011 & 0.0334 \\ 
  s(disease) & 1.0000 & 1.0000 & 0.5203 & 0.4786 \\ 
  s(discard) & 1.0000 & 1.0000 & 1.6726 & 0.2100 \\ 
  s(bycatch) & 1.0000 & 1.0000 & 0.9665 & 0.3367 \\ 
  s(mat\_pop) & 1.8919 & 1.9799 & 6.7981 & 0.0086 \\ 
  s(predation) & 1.0000 & 1.0000 & 2.9442 & 0.1009 \\ 
   \hline
\end{tabular}
\caption{GAM output for full model predicting mature mortality. Deviance explained =  72.04 \%} 
\label{tab.gam1}
\end{table}

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.1719 & 0.0105 & 16.4407 & $<$ 0.0001 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(disease) & 1.0000 & 1.0000 & 1.5639 & 0.2263 \\ 
  s(temperature) & 1.8837 & 1.9841 & 10.7801 & 0.0006 \\ 
  s(mat\_pop) & 2.0000 & 2.0000 & 13.2472 & 0.0002 \\ 
  s(imm\_pop) & 1.6241 & 1.8564 & 1.6100 & 0.1681 \\ 
  s(predation) & 1.0000 & 1.0000 & 0.1759 & 0.6796 \\ 
  s(bycatch) & 1.0000 & 1.0000 & 0.0006 & 0.9808 \\ 
  s(cannibalism) & 1.7298 & 1.9252 & 2.5843 & 0.1385 \\ 
   \hline
\end{tabular}
\caption{GAM output for full model predicting immature mortality. Deviance explained =  77.6 \%} 
\label{tab.gam2}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.6035 & 0.0467 & 12.9274 & $<$ 0.0001 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(temperature) & 1.9300 & 2.3189 & 4.6542 & 0.0183 \\ 
  s(mat\_pop) & 1.8887 & 1.9763 & 7.8385 & 0.0019 \\ 
   \hline
\end{tabular}
\caption{GAM output for trimmed model predicting mature mortality. Deviance explained =  62.16 \%} 
\label{tab.gam3}
\end{table}

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.1719 & 0.0121 & 14.1816 & $<$ 0.0001 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(temperature) & 1.7797 & 1.9506 & 9.5779 & 0.0006 \\ 
  s(mat\_pop) & 1.9813 & 1.9988 & 7.4968 & 0.0031 \\ 
   \hline
\end{tabular}
\caption{GAM output for trimmed model predicting immature mortality. Deviance explained =  59.5 \%} 
\label{tab.gam4}
\end{table}

## Sensitivities to model assumptions in GAMs
Modeling decisions are also necessarily made in the process of fitting GAMs and it is possible for these decisions to influence the outcome of an analysis. Within the context of our model, these decisions include the assumed error structure, the treatment of the uncertainty associated with the estimates of mortality, and the allowed shape of the smooths estimated in the GAMs. The following sensitivities address the implications of these modeling decisions on the outcome of our analysis.  
### Error structure
Impacts of assumptions about error structure were explored by assuming beta distributed data in the GAM and transforming the continuous total mortality rates to an exploitation rate ranging from 0 to 1. This transformation still resulted in mature population and temperature being the most important variables related to mortality, however the deviance explained decreased to 60% and 68% for mature and immature mortality, respectively, compared to 72% and 78% for the model presented in the main text. A potential shortcoming of the method presented in the main text is that predicted mortality could be less than zero. This was not the case in any of the model fittings, and would present a problem primarily if the model was extrapolated to data beyond the observed ranges. A potential fix to this issue is to log the response variable, but this resulted in unacceptable patterns in the residuals, so this model was not used.

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.0325 & 0.1850 & -0.1757 & 0.8623 \\ 
  disease & 0.0236 & 0.0162 & 1.4561 & 0.1607 \\ 
  imm\_pop & -0.0374 & 0.0174 & -2.1454 & 0.0442 \\ 
  predation & 0.0002 & 0.0169 & 0.0137 & 0.9892 \\ 
  bycatch & -0.0003 & 0.0147 & -0.0194 & 0.9847 \\ 
  cannibalism & -0.0090 & 0.0153 & -0.5887 & 0.5626 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(temperature) & 1.0005 & 1.0009 & 6.7766 & 0.0167 \\ 
  s(mat\_pop) & 1.7664 & 2.0738 & 3.5320 & 0.0466 \\ 
   \hline
\end{tabular}
\caption{GAM output for a shape constrained additive model predicting immature mortality. Deviance explained =  57.5 \%} 
\label{tab.gam5}
\end{table}

>>>>

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.6198 & 0.0451 & 13.7414 & $<$ 0.0001 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(temperature) & 2.1378 & 2.5245 & 4.7219 & 0.0169 \\ 
  s(disease) & 1.0000 & 1.0000 & 3.1106 & 0.0931 \\ 
  s(discard) & 1.0000 & 1.0000 & 0.7423 & 0.3991 \\ 
  s(bycatch) & 1.0000 & 1.0000 & 1.0154 & 0.3256 \\ 
  s(mat\_pop) & 1.8350 & 1.9593 & 4.1602 & 0.0252 \\ 
  s(predation) & 1.0000 & 1.0000 & 2.7661 & 0.1119 \\ 
   \hline
\end{tabular}
\caption{GAM output for full model predicting mature mortality. Deviance explained =  71.32 \%} 
\label{tab.gam1}
\end{table}

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.1741 & 0.0107 & 16.3429 & $<$ 0.0001 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(disease) & 1.0000 & 1.0000 & 0.7417 & 0.4004 \\ 
  s(temperature) & 1.8755 & 1.9825 & 11.4119 & 0.0005 \\ 
  s(mat\_pop) & 2.0000 & 2.0000 & 12.6704 & 0.0004 \\ 
  s(imm\_pop) & 1.6270 & 1.8596 & 1.7683 & 0.1424 \\ 
  s(predation) & 1.0000 & 1.0000 & 0.6671 & 0.4247 \\ 
  s(bycatch) & 1.0000 & 1.0000 & 0.0050 & 0.9445 \\ 
  s(cannibalism) & 1.7533 & 1.9368 & 2.8501 & 0.1134 \\ 
   \hline
\end{tabular}
\caption{GAM output for full model predicting immature mortality. Deviance explained =  78.43 \%} 
\label{tab.gam2}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.6198 & 0.0481 & 12.8854 & $<$ 0.0001 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(temperature) & 1.8741 & 2.2587 & 4.2891 & 0.0246 \\ 
  s(mat\_pop) & 1.8497 & 1.9631 & 6.7637 & 0.0036 \\ 
   \hline
\end{tabular}
\caption{GAM output for trimmed model predicting mature mortality. Deviance explained =  60.46 \%} 
\label{tab.gam3}
\end{table}

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.1741 & 0.0125 & 13.9569 & $<$ 0.0001 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(temperature) & 1.7403 & 1.9315 & 9.3351 & 0.0007 \\ 
  s(mat\_pop) & 1.9803 & 1.9985 & 6.6509 & 0.0056 \\ 
   \hline
\end{tabular}
\caption{GAM output for trimmed model predicting immature mortality. Deviance explained =  59.53 \%} 
\label{tab.gam4}
\end{table}

## Sensitivities to model assumptions in GAMs
Modeling decisions are also necessarily made in the process of fitting GAMs and it is possible for these decisions to influence the outcome of an analysis. Within the context of our model, these decisions include the assumed error structure, the treatment of the uncertainty associated with the estimates of mortality, and the allowed shape of the smooths estimated in the GAMs. The following sensitivities address the implications of these modeling decisions on the outcome of our analysis.

### Error structure
Impacts of assumptions about error structure were explored by assuming beta distributed data in the GAM and transforming the continuous total mortality rates to an exploitation rate ranging from 0 to 1. This transformation still resulted in mature population and temperature being the most important variables related to mortality, however the deviance explained decreased to 58% and 70% for mature and immature mortality, respectively, compared to 71% and 78% for the model presented in the main text. A potential shortcoming of the method presented in the main text is that predicted mortality could be less than zero. This was not the case in any of the model fittings, and would present a problem primarily if the model was extrapolated to data beyond the observed ranges. A potential fix to this issue is to log the response variable, but this resulted in unacceptable patterns in the residuals, so this model was not used.

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.0325 & 0.1850 & -0.1757 & 0.8623 \\ 
  disease & 0.0236 & 0.0162 & 1.4561 & 0.1607 \\ 
  imm\_pop & -0.0374 & 0.0174 & -2.1454 & 0.0442 \\ 
  predation & 0.0002 & 0.0169 & 0.0137 & 0.9892 \\ 
  bycatch & -0.0003 & 0.0147 & -0.0194 & 0.9847 \\ 
  cannibalism & -0.0090 & 0.0153 & -0.5887 & 0.5626 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(temperature) & 1.0005 & 1.0009 & 6.7766 & 0.0167 \\ 
  s(mat\_pop) & 1.7664 & 2.0738 & 3.5320 & 0.0466 \\ 
   \hline
\end{tabular}
\caption{GAM output for a shape constrained model predicting immature mortality. Deviance explained =  59.53 \%} 
\label{tab.gam5}
\end{table}
